---
title: "Modelling in R: A Chemostat"
author: "Thomas Petzoldt and Karline Soetaert"
description: "Tutorial: Equilibrium in a Chemostat"
date: "July-2021"
bibliography: chemostat.bib
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(deSolve)
library(rootSolve)
```



![](images/chemostat_d.svg){width=50%}

* $N$ nutrient, e.g. phosphorus (mmol P / m3)
* $P$ phytoplankton, carbon (mmol C / m3)

---

## Introduction

The example shows a chemostat model [cf. @Novick1950;@Herbert1956] by
example of heterotrophic or autotrophic organisms (bacteria, algae,
**P**), where growth is limited by a single nutrient (e.g. phosphorus,
**N**) via a Monod equation.

We use here a notation, compatible with the so-called NPZ and NPZD
models where **N** is a nutrient (nitrogen or phosphorus), **P** is
phytoplankton, **Z** zooplankton and (optionally) **D** is
detritus. Please don't get confused! Here **P** is phytoplankton and
not phosphorus, while the symbol for the nutrient **N** does not
necessarily mean that it is nitrogen.

The equations governing the dynamics of **P** and **N** are:
$$
\frac{dN}{dt}=-r \cdot \frac{N}{k_n+N}\cdot P + d \cdot(N_0-N)
$$
$$
\frac{dP}{dt}=r \cdot \frac{N}{k_n+N}\cdot P + d \cdot(0-P)
$$

where d is the dilution rate, and the nutrient and phytoplankton
concentration in the inflowing water is $N_0$ and $0$ respectively.

## Implementation

The system of differential equations is solved numerically with
solvers from package **deSolve** [@Soetaert2010a], while equilibria at
infinite time are estimated with package **rootSolve**
[@Soetaert2009]. The implementation follows the standard formulation
of these packages, see package documentation of **deSolve** and
**rootSolve**, and @Soetaert2010a or @Soetaert2010c for details.  The
code is written in the **R** programming language for statistical
computing [@RCore2018].

### The code

```{r, chemostat}
library("deSolve")
library("rootSolve")

chemostat <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
    p_growth <- r_pgrow * N / (k_n + N) * P

    dN_dt   <- - c_pn * p_growth + d * (N0 - N)
    dP_dt   <-          p_growth + d * (0 - P)
    list(c(dN_dt, dP_dt), p_growth = p_growth)
   })
}
parms <- c(
  r_pgrow = 0.5,   # phytoplankton growth parameter (1/d)
  k_n     = 1.0,   # half saturation constant, P (mmol/m3)
  c_pn    = 1/106, # stoichiometric conversion from phosphorus P to phyto C (P:C ratio)
  d       = 0.1,   # dilution rate 1/d
  N0      = 5      # phosphorus in inflow (mmol/m3)
)
```

### Dynamic simulation

A dynamic simulation can then be performed by function `ode`, using
the default algorithm `lsoda` with automatic integration step
size. The result (`out`) is a matrix-like object of class `deSolve`
that is supported by a generic (i.e. object oriented) plot function.

```{r, single, fig.height=3, fig.width=8}
times <- seq(from=0, to=40, length.out=100)  # simulation time (d)
y     <- c(N = 5, P = 1)                     # Nutrient (mmolP/m3) and Phytoplankton (mmolC/m3)
out   <- ode(y, times, chemostat, parms)
plot(out, mfrow=c(1, 3))
```

## Role of initial values

Let's play around with the model and change some values. First let's
modify the initial values `y <- c(N = 5, P = 1)`:

* Set `N`utrient and/or `P`hytoplankton to smaller or bigger values.
* Set `P`hytoplankton to zero.
* If necessary, change the time period.

```{r, chemostat-ex1, exercise=TRUE, exercise.lines = 28, fig.height=3, fig.width=8}
<<chemostat>>
times <- seq(from=0, to=40, length.out=100)  # simulation time (d)

y  <- c(N = .. , P = .. ) ##### Give values to N and P #####

out <- ode(y, times, chemostat, parms)
plot(out, mfrow=c(1, 3))
```

```{r chemostat-ex1-question, echo=FALSE}

ans <- "The equilibrium is independent on initial states, except if
  initial phytoplankton P is zero (nothing can grow without an
  inoculum). However, the initial state impacts the time it takes to
  reach equilibrium"

question("How do the initial values influence the final state?",
  answer("The final state depends on the initial states"),
  answer("The final state is independent on initial states, except `P=0`", correct = TRUE),
  answer("Initial state influence time until equilibrium", correct = TRUE),
  incorrect = ans,
  correct=ans,
  allow_retry = TRUE
)
```

## Role of the dilution rate $d$

Let's now observe the influence of the dilution rate parameter `d`:

* Set `d=0`, what does this mean?
* Set `d=0.5` what do you observe?
* Find a good value for `d` so that algae survive in the chemostat.

```{r, chemostat-ex2, exercise=TRUE, exercise.lines = 28, fig.height=3, fig.width=8}
<<chemostat>>
test_parms <- parms
test_parms["d"] <- 0.1
times <- seq(from=0, to=40, length.out=100)  # simulation time (d)
y     <- c(N = 5, P = 1)                     # Nutrient (mmolP/m3) and Phytoplankton (mmolC/m3)
out <- ode(y, times, chemostat, test_parms)
plot(out, mfrow=c(1, 3))
```

```{r chemostat-ex2-question, echo=FALSE}

ans <- "The chemostat can be run from `d=0` (batch) up to `d <
  r_pgrowth` (the maximum growth rate). If dilution exceeds growth,
  phytoplankton dies out."

question("What limits the range of the dilution rate so that algae survive in the chemostat?",
  answer("The dilution rate cannot be 0."),
  answer("A dilution rate of zero is equivalent to a batch.", correct = TRUE),
  answer("If the dilution rate exceeds max growth rate, phytoplankton dies out.", correct = TRUE),
  incorrect = ans,
  correct=ans,
  allow_retry = TRUE
)
```


## Solution: Scenarios with different patterns

Here we see some examples with different $d$. We see also that the
plot function allows to show several scenarios simultanaeously:

```{r, multiple, fig.height=4, fig.width=8}
times <- seq(from=0, to=40, length.out=100)  # simulation time (d)
y     <- c(N=5, P=1)
p1 <- p2 <- p3 <- parms
p1["d"] <- 0; p2["d"] <- 0.3; p3["d"] <- 0.5
out  <- ode(y, times, chemostat, parms)
out1 <- ode(y, times, chemostat, p1)
out2 <- ode(y, times, chemostat, p2)
out3 <- ode(y, times, chemostat, p3)

plot(out, out1, out2, out3, which=c("N", "P"))
```

## Steady State

In the sections above, we could see how the chemostat model approaches
an equilibrium. As the model is very simple, the equilibrium can be
calculated analytically by setting both state equations to zero and
then solving the system of equations. The solution can then be
plotted:


```{r analytical, fig.height=3, fig.width=8}
d <- seq(0, 0.6, length.out = 100)
r <- 0.5;  kp <- 0.5; Y = 106; N0 = 5
 
N <- d * kp / (r - d)
N <- ifelse(d > (r * N0)/(kp + N0), N0, N)
 
P <- Y * (N0 - N)
 
par(mfrow=c(1, 3))
plot(d, N,     type="l")
plot(d, P,     type="l")
plot(d, N * P, type="l")
```

Here we see that with increasing $d$ the equilibrium phytoplankton
concentration decreases and more nutrients are left over. The
production, however (N*P) approaches an optimum at high dilution
rates. If the dilution exceeds maximum growth, the chemostat dies out.

## Steady state solved numerically

To gain better understanding, we now want to estimate steady-state
numerically and recreate the relationship with dilution rate. For this
purpose, we may simulate the model for a long time until we see that
the solution does not change anymore (equilibrium) and then take the
last values.

But, as this is of course not very efficient, the equilibrium can also
approximated numerically with package **rootSolve**.

We first create a data frame, where the first column is a sequence of 
dilution rates within a sensible range. The other two columns are then reserved
for storing the results.

```{r, steadystate, exercise=TRUE, exercise.lines=30, fig.height=3, fig.width=8}
<<exercise>>
state <- data.frame(
  d = seq(0.01, 0.5, length.out = 100),
  N = 0,
  P = 0
)

## example how to calculate a single equilibrium

parms["d"] <- state$d[5]
times <- c(0, Inf)
out <- runsteady(y, times, chemostat, parms)
out$y

## place a loop here to fill the data frame .............

## and then outcomment the plot
#par(mfrow = c(1, 3))
#plot(N     ~ d, data = state, type = "l")
#plot(P     ~ d, data = state, type = "l")
#plot(N * P ~ d, data = state, type = "l")
```


```{r, steadystate-hint, fig.height=3, fig.width=8}
for (i in 1:nrow(state)) {
  parms["d"] <- state$d[i]
  times <- c(0, Inf)
  out <- runsteady(y, times, chemostat, parms)
  state[i, 2:3] <- out$y
}

par(mfrow = c(1, 3))
plot(N     ~ d, data = state, type = "l")
plot(P     ~ d, data = state, type = "l")
plot(N * P ~ d, data = state, type = "l")
``` 

We use a `for`-loop here for simplicity, even if list-based approaches
(e.g. `lapply`) may appear as more elegant for the intermediate and
advanced **R** user.

## Outlook

The model shown here shows some essential properties of a flow-through system,
used in biotechnology, wastewater treatment plants, or reservoirs and lakes. The
following tutorials will build up on this and extend the idea further.

### References

<div id="refs"></div>

----
