---
title: "Modelling in R: Introduction"
author: "Karline Soetaert and Thomas Petzoldt"
description: "This tutorial gives a brief introduction to the workshop. Watch the introductory video and try to answer a few questions."
date: "July-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

We recommend to watch the introductory video first:

![](https://youtu.be/L_zL_D1Qmjc)

If you prefer reading, you can find the slides under [this
link](https://dynamic-r.github.io/hacking-limnology/).



## About the authors (Thomas and Karline)

```{r background, echo=FALSE}

answertext <- "We are both biologists with a passion for mathematics;
  Thomas is a limnologist, Karline is a marine biologist. After her
  master in biology, Karline also studied computer sciences."

question("What is our scientific background?", type = "learnr_checkbox",
  answer("Mathematics"),
  answer("Statistics"),
  answer("Biology", correct=TRUE),
  answer("Biogeochemistry"),
  answer("Computer science", correct=TRUE),
  correct = answertext,
  incorrect = answertext
)

```

```{r Model_types, echo=FALSE}
answertext <- "We develop mechanistic models that start from
  principles of physics, chemistry, biology, ecology, and that are
  mathematically composed of differential equations. They are
  so-called “engineering type of models”, aiming to advance our
  understanding of natural systems and that allow prediction of these
  systems under various scenarios."

question("Which type of models do we typically use?", type = "learnr_checkbox",
  answer("Mechanistic models", correct=TRUE),
  answer("Statistical models"),
  answer("Black-box models"),
  answer("Qualitative models"),
  correct = answertext,
  incorrect = answertext
)
```

```{r why_modelling, echo=FALSE}

answertext <- "Modelling IS a lot of fun, but this is not our primary
  goal - we use models in addition to data gathering because it gives
  a different perspective of reality, it increases our understanding,
  allows to extrapolate, it may guide our research and may be used to
  generate hypotheses"

question("Why do we develop mechanistic models?", type = "learnr_checkbox",
  answer("they allow us to test our understanding of real systems", correct=TRUE),
  answer("they allow us to extrapolate to larger spatial and temporal scales", correct=TRUE),
  answer("they can guide empirical research", correct=TRUE),
  answer("they allow us to invalidate empirical research", correct=FALSE),
  answer("they allow us to test hypotheses", correct=FALSE),
  answer("they allow us to generate hypotheses", correct=TRUE),
  answer("because it is a lot of fun", correct=FALSE),
  incorrect = answertext,
  correct = answertext 
)
```

```{r programming_platform, echo=FALSE}

answertext <- "In this list, only R, matlab and python are programming
  languages, and R is the one that we use. Rmarkdown is a platform to
  merge R-code with text; Rstudio provides an interface that
  facilitates using R; deSolve is one of the many R-packages that we
  use"

question("Which programming language do we use to solve our models?", type = "learnr_checkbox",
  answer("Rstudio"),
  answer("Rmarkdown"),
  answer("R", correct=TRUE),
  answer("Matlab"),
  answer("Python"),
  answer("deSolve"),
  incorrect = answertext,
  correct = answertext 
)
```

```{r why_R, echo=FALSE}

answertext <- "We started using R around 2005-2010 because we thought
  it had potential, it was open-source, and supported by a growing
  community of R enthousiasts. Since then we created tools in R to
  make it suitable for our type of work. Even after all this time, R
  is still the best language to apply our type of models - it is ideal
  for teaching environmental modeling, as well as for more 'serious'
  scientific work. For the really complex models however, we resort to
  dedicated software, written in compiled code"

question("Why do we use R?", type = "learnr_checkbox",
  answer("it is faster than matlab or python"),
  answer("it is as good as matlab or python", correct= TRUE),
  answer("it is very well suited for post-processing", correct=TRUE),
  answer("it is an open-source community project", correct=TRUE),
  answer("the solvers have been optimised for mechanistic modeling", correct=TRUE),
  answer("it is easy to share models made in R", correct=TRUE),
  answer("we can run complex 3-D applications in R", correct=FALSE),
  incorrect = answertext,
  correct = answertext
)
```

## Model focus

```{r Model_focus, echo=FALSE}

answertext <- "Although we can do all the things listed above with
  mechanistic models, in environmental sciences, we restrict ourselves
  to exchanges of mass and energy and biogeochemistry/ecology relating
  to the environment"

question("What is the focus of environmental mechanistic models ?", type = "learnr_checkbox",
  answer("Exchange of mass", correct=TRUE),
  answer("Exchange of energy", correct=TRUE),
  answer("Biogeochemical processes in cells", correct=FALSE),
  answer("Impact of biogeochemical processes on the environment", correct=TRUE),
  answer("Predicting of human impacts on the environment", correct=TRUE),
  answer("Predicting the weather", correct=FALSE),
  answer("Predicting stock markets", correct=FALSE),
  incorrect = answertext,
  correct = answertext
)
```

## Finally


```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", 
  answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

### Please note

The feedback field is intended for future placement of this tutorial
on a server.  If running locally, it does currently not send its data
to us. We are nevertheless happy to get feedback in any form.
