---
title: "Modelling in R: From a Chemostat to NPZ with Phyto- and Zooplankton"
author: "Thomas Petzoldt and Karline Soetaert"
description: "Tutorial: extend a chemostat model towards an NPZ model with nutrient, phytoplankton and zooplankton"
date: "July-2021"
bibliography: chemostat.bib
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Introduction

The example shows how a chemostat model [cf. @Novick1950;@Herbert1956]
with heterotrophic or autotrophic organisms (bacteria, algae, $P$),
can be extended to a simple NPZ model with nutrients, phytoplankton
and zooplankton.

It builds on the chemostat model with two equations from the previous
tutorial.

Please don't get confused! Here **P** is phytoplankton and not
phosphorus, while the symbol for the nutrient **N** does not
necessarily mean that it is nitrogen.

## Implementation of the chemostat

The chemostat model is repeated here.  The model is written in the
**R** programming language for statistical computing [@RCore2018], and
solved numerically with solvers from package **deSolve**
[@Soetaert2010a]; equilibria are estimated with package **rootSolve**
[@Soetaert2009]. See @Soetaert2010a or @Soetaert2010c for details.

```{r, chemostat}
library("deSolve")
library("rootSolve")

chemostat <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
    p_growth <- r_pgrow * N / (k_n + N) * P

    dN_dt   <- -c_pn * p_growth + d * (N0 - N)
    dP_dt   <-         p_growth + d * (0 - P)
    list(c(dN_dt, dP_dt), p_growth = p_growth)
   })
}
parms <- c(
  r_pgrow = 0.5,   # phytoplankton growth parameter (1/d)
  k_n     = 1.0,   # half saturation constant, P (mmol/m3)
  c_pn    = 1/106, # stoichiometric conversion from phosphorus P to phyto C (P:C ratio)
  d       = 0.1,   # dilution rate (1/d)
  N0      = 5      # Phosphorus in inflow (mmol/m3)
)
times <- seq(from=0, to=40, by=0.1)  # simulation time (d)
y     <- c(N = 5, P = 1)             # Nutrient (mmolP/m3) N and Phytoplankton (mmolC/m3)

out <- ode(y, times, chemostat, parms) # solve model
plot(out, mfrow=c(1, 3))
```

## NPZ extension of the chemostat model

The chemostat equations can serve as a starting point for further
extensions, e.g. the production of specific substances in bioreactors,
or additional trophic levels. The equations are fundamental for the
description of other flow-through systems, up to waste water treatment
plants, or lakes.

Here we implement zooplankton as a third state variable and then test
the behavior of this more complex model.

Following the equations of $P$ and $N$ of the chemostat:

\begin{align}
\frac{dN}{dt} &= -r \cdot \frac{N}{k_n+N}\cdot P + d \cdot (N_0-N) \\
\frac{dP}{dt} &=  r \cdot \frac{N}{k_n+N}\cdot P + d \cdot (0-P)
\end{align}

we can add a 3rd equation for zooplankton $Z$ as:

\begin{align}
\frac{dZ}{dt} &= c_{asseff} \cdot c_{pn} \cdot r_{zgraz} \cdot \frac{P}{k_P + P} - r_{zloss} \cdot Z - d \cdot Z
\end{align}

with $c_{asseff}$ = assimilation efficiency, $r_{zgraz}$ = zooplankton grazing 
parameter, $r_{zloss}$ = zooplankton loss parameter, and $k_P$ = Monod constant 
of zooplankton growth on phytoplankton. Units and parameter values are found in the
code below.


Let's also assume, that nutrients from unused food $(1-c_{asseff})$ are recycled, 
so we can write the full system:

\begin{align}
    p_{growth}    &= r_{pgrow} \cdot \frac{N}{k_n + N}  \cdot P\\
    z_{grazing}   &= r_{zgraz}  \cdot \frac{P}{k_p + P} \cdot Z\\
    z_{growth}    &= c_{asseff} \cdot z_{grazing}\\
    n_{recycling} &= (1 - c_{asseff}) \cdot z_{grazing}\\
    \\
    \text{and the state equations:}\\
    \frac{dN}{dt}   &= c_{pn} \cdot (-p_{growth} + n_{recycling}) + d \cdot (N_0 - N)\\
    \frac{dP}{dt}   &= p_{growth} - z_{grazing}             - d \cdot P\\
    \frac{dZ}{dt}   &= z_{growth} - r_{zloss}  \cdot Z      - d \cdot Z
\end{align}


### Implementation of the NPZ model

```{r chemostat-npz, exercise=TRUE, exercise.eval=FALSE, exercise.lines = 25}
library("deSolve")
npz <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
 
    p_growth     <- r_pgrow * N / (k_n + N) * P
    
    z_grazing    <- ...
    z_growth     <- ...
    n_recycling  <- ...

    dN_dt   <- c_pn * (-p_growth + n_recycling) + d * (N0 - N)
    dP_dt   <- p_growth - z_grazing             + d * (0 - P)
    dZ_dt   <- ...
    
    list(c(dN_dt, dP_dt, ...))
   })
}
parms <- c(
  r_pgrow  = 0.5,   # phytoplankton growth parameter (1/d)
  r_zgraz  = 0.4,   # zooplankton grazing parameter (1/d)
  r_zloss  = 0.0,   # zooplankton loss parameter (1/d)
  
  k_n      = 0.5,   # Monod constant of phyto growth on nutrient (mmol/m3)
  k_p      = 100,   # Monod constant of zoo growth on phyto (mmol/m3)
  c_pn     = 1/106, # stoichiometric conversion from phosporus P to phyto C (P:C ratio)
  c_asseff = 0.3,   # zooplankton assimilation efficiency (-)

  d       = 0.1,    # dilution rate (1/d)
  N0      = 5       # P in inflow (mmol/m3)
)

times <- seq(0, 40, 0.1)        # (d)

# Nutrient in Phosphorus P; Phyto and Zoo in C and (mol/m3)
y  <- c(N=5, P=1, Z=10)

parms["d"] <- 0.1
out0 <- ode(y, times, npz, parms)
plot(out0, mfrow = c(1, 3))
```

```{r chemostat-npz-hint-1, exercise.lines = 1}
list(c(dN_dt, dP_dt, dZ_dt))
```

```{r chemostat-npz-hint-2, exercise.lines = 4}
    p_growth     <- r_pgrow * N / (k_n + N) * P
    z_grazing    <- r_zgraz  * P / (k_p + P) * Z
    z_growth     <- c_asseff * z_grazing
    n_recycling  <- (1 - c_asseff) * z_grazing
```

```{r chemostat-npz-hint-3,  exercise.lines = 3}
    dN_dt   <- c_pn * (-p_growth + n_recycling) + d * (N0 - N)
    dP_dt   <- p_growth - z_grazing             + d * (0 - P)
    dZ_dt   <- z_growth - r_zloss * Z           + d * (0 - Z)
```


```{r chemostat-npz-solution}
library("deSolve")
npz <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
 
    p_growth     <- r_pgrow * N / (k_n + N) * P
    
    z_grazing    <- r_zgraz  * P / (k_p + P) * Z
    z_growth     <- c_asseff * z_grazing
    n_recycling  <- (1 - c_asseff) * z_grazing

    dN_dt   <- c_pn * (-p_growth + n_recycling) + d * (N0 - N)
    dP_dt   <- p_growth - z_grazing             + d * (0 - P)
    dZ_dt   <- z_growth - r_zloss * Z           + d * (0 - Z)
    
    list(c(dN_dt, dP_dt, dZ_dt))
   })
}
parms <- c(
  r_pgrow  = 0.5,   # phytoplankton growth parameter (1/d)
  r_zgraz  = 0.4,   # zooplankton grazing parameter (1/d)
  r_zloss  = 0.0,   # zooplankton loss parameter (1/d)
  
  k_n      = 0.5,   # Monod constant of phyto growth on nutrient (mmol/m3)
  k_p      = 100,   # Monod constant of zoo growth on phyto (mmol/m3)
  c_pn     = 1/106, # stoichiometric conversion from phosporus P to phyto C (P:C ratio)
  c_asseff = 0.3,   # zooplankton assimilation efficiency (-)

  d       = 0.1,    # dilution rate (1/d)
  N0      = 5       # P in inflow (mmol/m3)
)

times <- seq(0, 40, 0.1)        # (d)

# Nutrient in Phosphorus P; Phyto and Zoo in C and (mol/m3)
y  <- c(N=5, P=1, Z=10)

parms["d"] <- 0.1
out0 <- ode(y, times, npz, parms)
plot(out0, mfrow = c(1, 3))
```

## Complete code

```{r chemostat-npz-code}
npz <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
 
    p_growth     <- r_pgrow * N / (k_n + N) * P
    
    z_grazing    <- r_zgraz  * P / (k_p + P) * Z
    z_growth     <- c_asseff * z_grazing
    n_recycling  <- (1 - c_asseff) * z_grazing

    dN_dt   <- c_pn * (-p_growth + n_recycling) + d * (N0 - N)
    dP_dt   <- p_growth - z_grazing             + d * (0 - P)
    dZ_dt   <- z_growth - r_zloss * Z           + d * (0 - Z)
    
    list(c(dN_dt, dP_dt, dZ_dt))
   })
}
parms <- c(
  r_pgrow  = 0.5,   # phytoplankton growth parameter (1/d)
  r_zgraz  = 0.4,   # zooplankton grazing parameter (1/d)
  r_zloss  = 0.0,   # zooplankton loss parameter (1/d)
  
  k_n      = 0.5,   # Monod constant of phyto growth on nutrient (mmol/m3)
  k_p      = 100,   # Monod constant of zoo growth on phyto (mmol/m3)
  c_pn     = 1/106, # stoichiometric conversion from phosporus P to phyto C (P:C ratio)
  c_asseff = 0.3,   # zooplankton assimilation efficiency (-)

  d       = 0.1,    # dilution rate (1/d)
  N0      = 5       # P in inflow (mmol/m3)
)

times <- seq(0, 40, 0.1)        # (d)

# Nutrient in Phosphorus P; Phyto and Zoo in C and (mol/m3)
y  <- c(N=5, P=1, Z=10)

parms["d"] <- 0.1
out0 <- ode(y, times, npz, parms)
plot(out0, mfrow = c(1, 3))
```


## Dynamic behavior

The model above is able to show various dynamic behavior, e.g. different 
equilibria and Lotka-Volterra cycles, depending on the parameter values.

### Exercise

Adapt the model parameters and try to reproduce the following dynamic patterns:


* scenario 0: equilibrium, algae extinct
* scenario 1: Lotka-Volterra cycle
* scenario 2: equilibrium, zooplankton extinct
* scenario 3: equilibrium, coexistence
* scenario 4: equilibrium, both extinct
* scenario 5: damped oscillation

Start to change the parameter $d$ only and if time allows play also with other parameters.

```{r npz-scenario-exercise, exercise = TRUE, exercise.lines = 20, exercise.setup="chemostat-npz-code", ig.height=3, fig.width=8}
times <- seq(0, 1000)
parms["d"] <- 0.0
out <- ode(y, times, npz, parms)
plot(out, mfrow = c(1, 3))
```


## Dynamic bevavior: Solutions

```{r npz-scenario0, fig.height=3, fig.width=8}
times <- seq(0, 1000)
parms["d"] <- 0.0
out0 <- ode(y, times, npz, parms)
plot(out0, mfrow = c(1, 3))
```

```{r npz-scenario1, fig.height=3, fig.width=8}
parms["d"] <- 0.2
out1 <- ode(y, times, npz, parms)
plot(out1, mfrow=c(1, 3))
```

```{r npz-scenario2, fig.height=3, fig.width=8}
parms["d"] <- 0.07
out2 <- ode(y, times, npz, parms)
plot(out2, mfrow=c(1, 3))
```

```{r npz-scenario3, fig.height=3, fig.width=8}
times <- seq(0, 1000)
parms["d"] <- 0.09
out3 <- ode(y, times, npz, parms)
plot(out3, mfrow=c(1, 3))
```


```{r npz-scenario4, fig.height=3, fig.width=8}
parms["d"] <- 0.5
out4 <- ode(y, times, npz, parms)
plot(out4, mfrow=c(1, 3))
```

```{r npz-damped, fig.height=3, fig.width=8}
# scenario X: damped Lotka-Volterra cycle
parms["d"] <- 0.081
out5 <- ode(y, times, npz, parms)
plot(out5, mfrow=c(1, 3))
```


```{r npz-all, fig.height=3, fig.width=8}
plot(out0, out1, out2, out3, out4, out5, mfrow=c(1,3), lty=1)
legend("topright",
  legend=c("EQ, No Algae",
           "Lotka-Volterra",
	         "EQ, No Zoo", 
           "EQ, coexistence", 
           "EQ. both extinct", 
           "Damped oscillation"), 
  lty=1, col=1:6, cex=0.5)
```

## Outlook

The example shows that even such a simple model can show surprisingly complex
behavior. It can therefore serve as a starting point in different directions,
practically by extending the model and making it more realistic, and theoretically
by analysing its behavior more systematically by, for example, stability and 
bifurcation analysis.


### References

<div id="refs"></div>

----
