---
title: "1D NPZ Model in a River"
author: "Karline Soetaert and Thomas Petzoldt"
description: "This tutorial explains elements of a 1D Model in a River."
date: "July-2021"
bibliography: aubergine.bib
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{1D NPZ Model in a River}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
library("ReacTran")
library("rootSolve")
```

## Introduction

The NPZ model implemented in the previous tutorial can easily be extended from 0D 
to 1D, using the R-package **ReacTran** [@Soetaert2012ReacTran]. Basically, 

* the spatial domain is subdivided in a number of grid cells (function `setup.grid.1D`), and 
* the state equations are vectorized so that they are solved for all grid cells in parallel.
* transport (advective flow and mixing) is performed by function `tran.1D`
* the numerical methods `ode.1D` and `steady.1D` from **deSolve** and **rootSolve** are optimised to solve one-dimensional transport problems.\footnote{`ode.1D` and `steady.1D` know the structure of the Jacobian to be 'banded' (as only adjacent grid cells depend on each other), making it possible to efficiently estimate the Jacobian}

## From 0D to 1D

We start with the 0D chemostat model from previous tutorial that describes the 
dynamics of Nutrients and Phytoplankton:

```{r, chemostatNP}
library (deSolve)
chemostatNP <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
    p_growth <- r_pgrow * N / (k_n + N) * P
    
  # mass balance = process rates   + transport
    dN_dt   <-  - c_pn * p_growth + d*(N0-N)
    dP_dt   <-           p_growth + d*(0 -P)
    list(c(dN_dt, dP_dt))
   })
}
parms<- c(r_pgrow=0.5, k_n=1.0, c_pn=1/106, d=0.1, N0=5)
y    <- c(N = 5, P = 1)  
std  <- steady(y, times=c(0,Inf), func=chemostatNP, parms=parms,
                method="runsteady") # solve model
std$y
```

The code below shows how this zero-dimensional NP model is extended to 1D.

```{r river-np, fig.height=4, fig.width=7}
library(ReacTran)                                               ## CHANGED ##

riverNP <- function(time, y, parms) {
  with(as.list(parms), {    # was:  as.list(c(y, parms))        ## CHANGED ##
   # state vectors need to be extracted from y                  ## NEW ##
    N <- y[        1 :(1*nbox)]   
    P <- y[(1*nbox+1):(2*nbox)]   

    p_growth  <- r_pgrow * N / (k_n + N) * P

   # transport: C.up=upstream conc, v=flow velocity             ## NEW ##
    n_transport <- tran.1D(C=N, C.up=N0, v=v, dx=Grid)  
    p_transport <- tran.1D(C=P, C.up=P0, v=v, dx=Grid)

   # mass balance = process rates + transport
    dN_dt <- c_pn * (-p_growth)  + n_transport$dC               ## CHANGED ##
    dP_dt <-          p_growth   + p_transport$dC

    list(c(dN_dt, dP_dt))
  })
}
parms <- c(
  r_pgrow = 0.5,   # phytoplankton growth parameter (1/d)
  k_n     = 0.5,   # Monod ct of phyto growth on nutrient (mmol/m3)
  k_p     = 100,   # Monod ct of zoo grazing on phyto (mmol/m3)
  c_pn    = 1/106, # conversion from phosporus to phyto (P:C)
  v       = 10000, # flow velocity (m/d)                        ## NEW ##
  N0      = 500,   # nutrient (PO4) in inflow (mmolP/m3)    
  P0      = 10     # algae in inflow (mmolC/m3)
)

# morphology: divide the river into 500 boxes                   ## NEW ##
Length <- 500e3  # length of the river (m)
nbox   <- 500    # number of 1D grid cells
Grid   <- setup.grid.1D(L = Length, N = nbox)

# Phyto as C and Phosphorus P (mol/m3) - each nbox long         ## CHANGED ##
N_init  <- rep(5, times=nbox)  # mmolP/m3
P_init  <- rep(1, times=nbox) # mmolC/m3
y_init  <- c(N_init, P_init)

std <- steady.1D(y_init, times=c(0,Inf), func=riverNP, parms=parms,
                 nspec=2, names=c("N","P"),                     ## NEW ##
                 method="runsteady")
plot(std, mfrow=c(1,2), grid=Grid$x.mid/1000,                   ## CHANGED ## 
     lty=1, lwd=2, xlab="km")
```

## The 1D NPZ model

You now get the extended NPZ model in the chemostat:

```{r, chemostatNPZ}
library (deSolve)
chemostatNPZ <- function(time, y, parms) {
  with(as.list(c(y, parms)), {
    p_growth  <- r_pgrow * N / (k_n + N) * P
    z_grazing <- r_zgraz * P / (k_p + P) * Z
    z_loss    <- r_zloss * Z
  # mass balance = process rates   + transport
    dN_dt <- c_pn*(-p_growth + (1 - asseff) * z_grazing + z_loss) + d * (N0 - N)
    dP_dt <-        p_growth -                z_grazing           + d * (0 - P)
    dZ_dt <-                        asseff  * z_grazing - z_loss  + d * (0 - Z)
    list(c(dN_dt, dP_dt, dZ_dt))
   })
}
parms<- c(r_pgrow=0.5, r_zgraz=0.4, r_zloss=0.01, k_n=0.5, 
          k_p=100, c_pn=1/106, asseff=0.3, d=0.11, N0=500)
y   <- c(N=5, P=1, Z=1)  
std <- steady(y, times=c(0,Inf), func=chemostatNPZ, parms=parms, method="runsteady") 
```

Your task is to extend the NPriver model to include the Z component:
```{r riverNPZ, eval=FALSE, fig.height=4, fig.width=7}
library(ReacTran)           

riverNPZ <- function(time, y, parms) {
  with(as.list(parms), {    
   # state vectors need to be extracted from y                 
    N <- y[        1 :(1*nbox)]   
    P <- y[(1*nbox+1):(2*nbox)]   
    Z <- ..                                                     ## CHANGE ##
    
    p_growth  <- r_pgrow * N / (k_n + N) * P
    z_grazing <- ...                                            ## ADD ##
    z_loss    <- ...                                            ## ADD ##

   # transport: C.up=upstream conc, v=flow velocity              
    n_transport <- tran.1D(C=N, C.up=N0, v=v, dx=Grid)  
    p_transport <- tran.1D(C=P, C.up=P0, v=v, dx=Grid)
    z_transport <- ..                                           ## ADD ##
    
   # mass balance = process rates + transport
    dN_dt <- c_pn * (-p_growth)  + n_transport$dC               ## CHANGE ##
    dP_dt <-          p_growth   + p_transport$dC               ## CHANGE ##
    dZ_dt <- ..                                                 ## ADD ##

    list(c(dN_dt, dP_dt, ..))                                   ## CHANGE ##
  })
}
parms <- c(
  r_pgrow  = 0.5,   # phytoplankton growth parameter (1/d)
  r_zgraz  = 0.4,   # zooplankton ingestion parameter (1/d)
  r_zloss  = 0.01,  # zooplankton loss parameter (1/d)

  k_n      = 0.5,   # Monod ct of phyto growth on nutrient (mmol/m3)
  k_p      = 100,   # Monod ct of zoo grazing on phyto (mmol/m3)
  c_pn     = 1/106, # conversion from phosporus P to phyto C (P:C)
  asseff   = 0.3,   # assimilation efficiency of zoo (-)

  v        = 10000, # flow velocity (m/d)
  N0       = 500,   # nutrient (PO4) in inflow (mmolP/m3)
  P0       = 10,    # algae in inflow (mmolC/m3)
  Z0       = 1      # zooplankton in inflow (mmolC/m3)
)

# river morphology
Length <- 500e3  # length of the river (m)
nbox   <- 500    # number of 1D grid cells
Grid   <- setup.grid.1D(L = Length, N = nbox)

# Zoo and Phyto as C and Phosphorus P (mol/m3)
N_init  <- rep(500, times=nbox) # mmolP/m3
P_init  <- rep(100, times=nbox) # mmolC/m3
Z_init  <- ..                                                   ## ADD ##
y_init  <- c(N_init, P_init, ..)                                ## ADD ##

std <- steady.1D(y=y_init, time=0, func=riverNPZ, parms=parms,
                 nspec=2, names=c("N","P"),                     ## CHANGE ##
                 method="runsteady")
plot(std, mfrow=c(1,3), grid=Grid$x.mid/1000, xlab="km")
```

## Solution

The code of the solution is not included in the document, but can be seen in the 
source code.

```{r riverNPZ-solution, include=FALSE, fig.height=3, fig.width=7}
library(ReacTran)

riverNPZ <- function(time, y, parms) {
  with(as.list(parms), {
    N <- y[       1  :   nbox ]
    P <- y[(  nbox+1):(2*nbox)]
    Z <- y[(2*nbox+1):(3*nbox)]                            ## ADDED ##

    p_growth  <- r_pgrow * N / (k_n + N) * P
    z_grazing <- r_zgraz * P / (k_p + P) * Z               ## ADDED ##
    z_loss    <- r_zloss * Z                               ## ADDED ##   

    n_transport <- tran.1D(C=N, C.up=N0, v=v, dx=Grid)
    p_transport <- tran.1D(C=P, C.up=P0, v=v, dx=Grid)
    z_transport <- tran.1D(C=Z, C.up=Z0, v=v, dx=Grid)     ## ADDED ##

    dN_dt <- c_pn*(-p_growth + (1-asseff)*z_grazing + z_loss) + n_transport$dC
    dP_dt <-        p_growth -            z_grazing           + p_transport$dC
    dZ_dt <-                 +    asseff *z_grazing - z_loss  + z_transport$dC

    list(c(dN_dt, dP_dt, dZ_dt))
  })
}
parms <- c(
  r_pgrow  = 0.5,   # phytoplankton growth parameter (1/d)
  r_zgraz  = 0.4,   # zooplankton ingestion parameter (1/d)
  r_zloss  = 0.01,  # zooplankton loss parameter (1/d)

  k_n      = 0.5,   # Monod constant of phyto growth on nutrient (mmol/m3)
  k_p      = 100,   # Monod constant of zoo grazing on phyto (mmol/m3)
  c_pn     = 1/106, # stoichiometric conversion from phosporus P to phyto C (P:C)
  asseff   = 0.3,   # assimilation efficiency of zoo (-)

  v        = 10000, # flow velocity (m/d)
  N0       = 500,   # nutrient (PO4) in inflow (mmolP/m3)
  P0       = 10,    # algae in inflow (mmolC/m3)
  Z0       = 1      # zooplankton in inflow (mmolC/m3)
)

# river morphology
Length <- 500e3  # length of the river (m)
nbox   <- 500    # number of 1D grid cells
Grid   <- setup.grid.1D(L = Length, N = nbox)

# Zoo and Phyto as C and Phosphorus P (mol/m3)
N_init  <- rep(500, times=nbox) # mmolP/m3
P_init  <- rep(100, times=nbox) # mmolC/m3
Z_init  <- rep(10,  times=nbox) # mmolC/m3
y_init  <- c(N_init, P_init, Z_init)
```

Now the steady state can be found with function `steady.1D` from package 
**rootSolve**:

```{r riverNPZ-solution-plot, fig.height=3, fig.width=7}
std <- steady.1D(y=y_init, time=0, func=riverNPZ, parms=parms,
                 nspec=3, names=c("N","P","Z"), method="runsteady")
plot(std, mfrow=c(1,3))
```

## The 1D NPZ model - scenarios

The river model with nutrient, phytoplankton and zooplankton dynamics is now run
with different flow rates. Flow rates are varied from 1 km /day up to 10 km per day:

```{r riverNPZscenario, fig.height=8, fig.width=7}
parms0 <- parms1 <- parms2 <- parms
parms0["v"] <- 1000  # m/d
std0 <- steady.1D(y=y_init, time=0, func=riverNPZ, parms=parms0,
                  nspec=3, names=c("N","P","Z"), method="runsteady")

parms1["v"] <- 5000  # m/d
std1 <- steady.1D(y=y_init, time=0, func=riverNPZ, parms=parms1,
                  nspec=3, names=c("N","P","Z"), method="runsteady")

parms2["v"] <- 10000 # m/d
std2 <- steady.1D(y=y_init, time=0, func=riverNPZ, parms=parms2,
                  nspec=3, names=c("N","P","Z"), method="runsteady")

plot(std0, std1, std2, mfrow=c(3,1), grid=Grid$x.mid/1000, 
     lty=1, lwd=2, xlab="km")
legend("top", title="v=", legend=c("1km/d","5km/d","10km/d"), 
       col=1:3, lty=1)
```

## Outlook

The 1D-NPZ model shows how straightforward transport can be implemented. We used 
the "method of lines" where space is discretized in fixed steps while time is
solved with standard solvers in variable time steps. This works for many problems,
even in 2D and 3D [cf. @Soetaert2008;@Soetaert2012], but there are of course also 
specialized cases where other methods are preferred.

The model described here can be seen as a starting point for more complex 
applications, e.g. with seasonal (or diurnal) forcing, variable river geometry,
vertical exchange in the weater column of lakes or in the sediment and of course 
also for completely different problems in other fields.

## References

<div id="refs"></div>

----
