---
title: "Modelling in R: An NPZD Model"
author: "Karline Soetaert and Thomas Petzoldt"
description: "This tutorial explains elements of an NPZD model."
date: "July-2021"
bibliography: bib.bib
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(knitr)
```


## Introduction

**Note:** Though the following example is somewhat fragmentary, it is included 
for two reasons:

* to give a practical example how models can be driven with forcing data
* to point to the "Practical Guide to Ecological Modelling"-book with more related
examples.

The model is a minimally modified version taken from [@Soetaert2008] and 
it is also found online at [the book homepage](https://www.springer.com/gp/book/9781402086236).

The notation differs somewhat from what was used in the previous examples, and 
allows direct comparison with the book.

The model illustrates two things:

1. how to extend the NPZ model with a 4th state variable for detritus
2. How to implement seasonal forcing.

## Forcing functions as part of the model

In the version given below,
photosynthetic radiation (PAR) is given by a forcing function directly in the model code,
but in practice, input data are often given as tables.


```{r npzd-sin}
library("deSolve")
NPZD <-  function(time, state, parameters){
 with(as.list(c(state, parameters)),{
    # Light, a sine function; 50% of light is PAR
    PAR            <- 0.5 * (540 + 440 * sin(2 * pi * time/365 - 1.4))

    Nuptake        <- maxUptake * PAR/(PAR + ksPAR) * DIN/(DIN + ksDIN) * PHYTO

    Grazing        <- maxGrazing * PHYTO/(PHYTO + ksGrazing) * ZOO
    Faeces         <- pFaeces * Grazing
    Excretion      <- excretionRate * ZOO
    Mortality      <- mortalityRate * ZOO * ZOO
    Mineralisation <- mineralisationRate * DETRITUS
    Chlorophyll    <- chlNratio * PHYTO
    TotalN         <- PHYTO + ZOO + DETRITUS + DIN

    dPHYTO    <- Nuptake - Grazing
    dZOO      <- Grazing - Faeces - Excretion - Mortality
    dDETRITUS <- Mortality - Mineralisation + Faeces
    dDIN      <- Mineralisation + Excretion - Nuptake

    # the output, packed as a list
    list(c(dPHYTO, dZOO, dDETRITUS, dDIN),       # rate of change
        c(Chlorophyll = Chlorophyll, PAR = PAR)) # extra outputs
  })
}

parms = c(maxUptake          = 1.0,       # /day
          ksPAR              = 140,       # µEinst/m2/s
          ksDIN              = 0.5,       # mmolN/m3
          maxGrazing         = 1.0,       # /day
          ksGrazing          = 1.0,       # mmolN/m3
          pFaeces            = 0.3,       # -
          excretionRate      = 0.1,       # /day
          mortalityRate      = 0.4,       # /(mmolN/m3)/day
          mineralisationRate = 0.1,       # /day
          chlNratio          = 1          # mgChl/mmolN
)
times  = seq(from = 0, to = 365, by = 1)  # time step is days
init   =  c(PHYTO   = 1,                  # state variable initial conditions, units mmolN/m3
            ZOO     = 0.1,
            DETRITUS= 5.0,
            DIN     = 5.0)
```

```{r npzd-ode, fig.height=8}
out <- ode(y=init, times=times, func=NPZD, parms=parms)
plot(out)
```

## Tabular interpolation

Let's assume we have a table of PAR data:

```{r npzd-inputs-create}
time <- round(seq(1, 365, length.out=25), 0)
PAR  <- round(0.5 * (540 + 440 * sin(2 * pi * time/365 - 1.4)), 0)
inputs <- data.frame(
  time = time,
  PAR  = PAR
)
```

```{r npzd-inputs-show}
kable(inputs)
```




```{r npzd-approxfun}
library("deSolve")
NPZD <-  function(time, state, parameters){
 with(as.list(c(state, parameters)),{
    # Light, a sine function; 50% of light is PAR
    #PAR            <- 0.5 * (540 + 440 * sin(2 * pi * time/365 - 1.4))
    PAR <- parfunction(time)

    Nuptake        <- maxUptake * PAR/(PAR + ksPAR) * DIN/(DIN + ksDIN) * PHYTO

    Grazing        <- maxGrazing * PHYTO/(PHYTO + ksGrazing) * ZOO
    Faeces         <- pFaeces * Grazing
    Excretion      <- excretionRate * ZOO
    Mortality      <- mortalityRate * ZOO * ZOO
    Mineralisation <- mineralisationRate * DETRITUS
    Chlorophyll    <- chlNratio * PHYTO
    TotalN         <- PHYTO + ZOO + DETRITUS + DIN

    dPHYTO    <- Nuptake - Grazing
    dZOO      <- Grazing - Faeces - Excretion - Mortality
    dDETRITUS <- Mortality - Mineralisation + Faeces
    dDIN      <- Mineralisation + Excretion - Nuptake

    # the output, packed as a list
    list(c(dPHYTO, dZOO, dDETRITUS, dDIN),       # rate of change
        c(Chlorophyll = Chlorophyll, PAR = PAR)) # extra outputs
  })
}

parms = c(maxUptake          = 1.0,       # /day
          ksPAR              = 140,       # µEinst/m2/s
          ksDIN              = 0.5,       # mmolN/m3
          maxGrazing         = 1.0,       # /day
          ksGrazing          = 1.0,       # mmolN/m3
          pFaeces            = 0.3,       # -
          excretionRate      = 0.1,       # /day
          mortalityRate      = 0.4,       # /(mmolN/m3)/day
          mineralisationRate = 0.1,       # /day
          chlNratio          = 1          # mgChl/mmolN
)
times  = seq(from = 0, to = 365, by = 1)  # time step is days
init   =  c(PHYTO   = 1,                  # state variable initial conditions, units mmolN/m3
            ZOO     = 0.1,
            DETRITUS= 5.0,
            DIN     = 5.0)
```

Now lets define an interpolation funtion. Here we use R's `approxfun`, that is a 
function that returns a function. It is technically callled a "closure", because
the data used during creation become part of it.

It can then be called with a given time and returns the interpolated dependent 
variable (PAR):

```{r npzd-ode-approxfun, fig.height=8}
parfunction <- approxfun(x=inputs$time, y=inputs$PAR, rule=2)
parfunction(56)
```


```{r npzd-ode2, fig.height=8}
out <- ode(y=init, times=times, func=NPZD, parms=parms)
plot(out)
```

In the code above, `parfunction` is passed to the model function as a global 
variable. If you, however like to pass it explicitly to the model, add an additional 
parameter to the function, e.g.:

```
## in the model function
NPZD <-  function(time, state, parameters, parfunction) {

...

}

## and in the call
out <- ode(y=init, times=times, func=NPZD, parms=parms, parfunction=parfunction)
```

## Outlook

More information about the model is found in [@Soetaert2008], and how to use 
forcings in the "forcings" help (`help("forcings")`) page of 
the **deSolve** package and the tutorial at [https://tpetzoldt.github.io](https://tpetzoldt.github.io/deSolve-forcing/deSolve-forcing.html).



### References

<div id="refs"></div>

----
